---
title: "Untitled"
author: "Adedamola Ogundipe (R)"
date: "`r Sys.Date()`"
output: html_document
---

### LOAD LIBRARY AND DATASET
```{r Library and Dataset}
library(readr)
data <- read_csv("/Users/m1/Desktop/OGUNDIPE/PAU/Notes/Datasets/Diabetes Dataset.csv")
```

## DATA CLEANING AND PREPROCESSING
## Summary of Data, Data Manipulation and cleaning
Inspection of dataset for 
-Find missing values
-Find outliers
-Other inconsistencies

## Specifications and Quick glimpse of data
```{r quick glance of data}
library(dplyr)
spec(data)
glimpse(data)
```

## Find Missing Values
```{r finiding missing data}
sum(is.na(data))
```

## Detect and remove Outliers in Major Features by Filtering

```{r Outlier detection and removal}
par(mfrow = c(1, 4))
boxplot(data$BMI, main="BMI Boxplot")
boxplot(data$SBP, main="SBP Boxplot")
boxplot(data$DBP, main="DBP Boxplot")
boxplot(data$Chol, main="Chol Boxplot")
data_updated <- data[data$BMI < 50 & data$BMI > 15 & data$SBP < 240 & data$DBP > 50 & data$Chol < 10, ]
par(mfrow = c(1, 4))
boxplot(data_updated$BMI, main="BMI Boxplot")
boxplot(data_updated$SBP, main="SBP Boxplot")
boxplot(data_updated$DBP, main="DBP Boxplot")
boxplot(data_updated$Chol, main="Chol Boxplot")
```

## Feature Engineering

```{r feature engineering}
data_updated$MAP <- with(data_updated, DBP + (1/3) * (SBP - DBP))
data_updated$BMI_Category <- cut(data_updated$BMI, 
                               breaks=c(-Inf, 18.5, 25, 30, Inf), 
                               labels=c("Underweight", "Normal", "Overweight", "Obese"))
table(data_updated$BMI_Category)
```

## Cross Check Data
```{r Another quick glance at data post-processing}
str(data_updated)
```

# Visualizations
```{r visualizations}
library(ggplot2)
ggplot(data_updated, aes(x=BMI)) + geom_histogram(binwidth=1, fill="blue", color="black") + 
  ggtitle("BMI Distribution")
ggplot(data_updated, aes(x=Age)) + 
  geom_histogram(binwidth=1, fill="blue", color="black") + 
  ggtitle("Age Distribution")
ggplot(data_updated, aes(x=SBP)) + 
  geom_histogram(binwidth=1, fill="green", color="black") + 
  ggtitle("Systolic Blood Pressure (SBP) Distribution")
ggplot(data_updated, aes(x=DBP)) + 
  geom_histogram(binwidth=1, fill="red", color="black") + 
  ggtitle("Diastolic Blood Pressure (DBP) Distribution")

ggplot(data_updated, aes(x=Chol)) + 
  geom_histogram(binwidth=1, fill="purple", color="black") + 
  ggtitle("Cholesterol Distribution")
ggplot(data_updated, aes(x=Age, y=BMI, color=factor(Diabetes))) + 
  geom_point() + 
  ggtitle("BMI vs. Age by Diabetes Status")
ggplot(data_updated, aes(x = as.factor(Diabetes), y = Age)) + 
  geom_boxplot(fill = "cyan", color = "black") +
  theme_minimal() +
  ggtitle("Age Distribution by Diabetes Status")
```

## Converting Target Variable from numerical to factor

```{r conversion of target variable to factor }
data_updated$Diabetes <- as.factor(data_updated$Diabetes)
levels(data_updated$Diabetes) <- c("0", "1")